---
title: "CIND 820 - Big Data Analytics Project"
author: 
output:
  html_document: default
  word_document: default
  pdf_document: default
---
<center> <h1> Initial Analysis </h1> </center>
<center> <h3> Jeffrey Fitzpatrick </h3> </center>
<center> <h3> 500728133 </h3> </center>
---

## 

Room Occupancy Estimation is a data set for estimating the precise number of occupants in a room using
non-intrusive environmental sensors like temperature, light, sound, CO2 and PIR.The data set is available
from the UC Irvine Machine Learning Repository (https://archive.ics.uci.edu/dataset/864/room+occupancy+estimation) in CSV format.

## Univariate Analysis

Create a data frame of the Room Occupancy Estimation dataset. Display the names of the columns.

```{r}
# Import the Occupancy_Estimation.csv file
occupancy_estimation <- read.csv("Occupancy_Estimation.csv")

# Display the names of the columns
names(occupancy_estimation)
```
Rename the Room_Occupancy_Count column

```{r}
names(occupancy_estimation)[19] <- "Occupancy_Count"
names(occupancy_estimation)
```


Check the data types of the attributes.

```{r}
# Check the data types of the attributes
str(occupancy_estimation)
```
Run a summary statistics of the dataset.

```{r}
# Run a summary statistics of the numeric variables
summary(occupancy_estimation[c("S1_Temp", "S2_Temp","S3_Temp", "S4_Temp",
                               "S1_Light", "S2_Light","S3_Light", "S4_Light",
                               "S1_Sound", "S2_Sound","S3_Sound", "S4_Sound",
                               "S5_CO2", "S5_CO2_Slope")])
```
Check for missing values in the data set.

```{r}
# Check the missing values in the data set
sum(is.na(occupancy_estimation))
```
There are no missing values in the data set.

The Date variable is seen as a character string. Make it a date.
Check the data types of the attributes.

```{r}
# Convert Date variable to a date
occupancy_estimation$Date <- as.Date(occupancy_estimation$Date, format="%Y/%m/%d")

# Check the data types of the attributes
str(occupancy_estimation)
```
Use a histogram to assess normality of the Temperature variables.

```{r}
par(mfrow=c(2,2))

# Check if the distribution of S1 Temperature is normally distributed
hist(occupancy_estimation$S1_Temp, 
     xlab="Temperature (Celsius)", 
     main="Histogram of Sensor 1 Temperatures",
     col="skyblue",
     border="black",
     breaks=8)
grid()

# Check if the distribution of S2 Temperature is normally distributed
hist(occupancy_estimation$S2_Temp, 
     xlab="Temperature (Celsius)", 
     main="Histogram of Sensor 2 Temperatures",
     col="skyblue",
     border="black",
     breaks=8)
grid()

# Check if the distribution of S3 Temperature is normally distributed
hist(occupancy_estimation$S3_Temp, 
     xlab="Temperature (Celsius)", 
     main="Histogram of Sensor 3 Temperatures",
     col="skyblue",
     border="black",
     breaks=8)
grid()

# Check if the distribution of S4 Temperature is normally distributed
hist(occupancy_estimation$S4_Temp, 
     xlab="Temperature (Celsius)", 
     main="Histogram of Sensor 4 Temperatures",
     col="skyblue",
     border="black",
     breaks=8)
grid()

par(mfrow=c(1,1))
```
```{r}
# Normality test for sensor 4 temperature. Sample size must be between 3 and 5000.
temp <- subset(occupancy_estimation, Date=="2017-12-22" | Date=="2017-12-23" | Date=="2017-01-10") # Days with occupants

shapiro_test <- shapiro.test(temp$S4_Temp)
print(shapiro_test)

if (shapiro_test$p.value > 0.05) {
  print("Temperature data for sensor 4 is normally distributed")
} else {
  print("Temperature data for sensor 4 is not normally distributed")
}

```
Create a boxpot which shows the Temperature values of each sensor. Use different colors for each sensor.

```{r}
# Create the boxplot
boxplot(occupancy_estimation$S1_Temp, occupancy_estimation$S2_Temp, occupancy_estimation$S3_Temp, 
        occupancy_estimation$S4_Temp, 
        names=c("Sensor 1", "Sensor 2", "Sensor 3", "Sensor 4"),
        col=c("green","magenta","orange","red"),
        ylab = "Temperature (Celsius)", main="Boxplot of Temperature Values by Sensor" )
grid()
```

Use a histogram to assess normality of the Light variables.

```{r}
par(mfrow=c(2,2))

# Check if the distribution of S1 Light is normally distributed
hist(occupancy_estimation$S1_Light, 
     xlab="Light (Lux)", 
     main="Histogram of Sensor 1 Light Values",
     col="skyblue",
     border="black",
     breaks=8)
grid()

# Check if the distribution of S2 Light is normally distributed
hist(occupancy_estimation$S2_Light, 
     xlab="Light (Lux)", 
     main="Histogram of Sensor 2 Light Values",
     col="skyblue",
     border="black",
     breaks=8)
grid()

# Check if the distribution of S3 Light is normally distributed
hist(occupancy_estimation$S3_Light, 
     xlab="Light (Lux)", 
     main="Histogram of Sensor 3 Light Values",
     col="skyblue",
     border="black",
     breaks=8)
grid()

# Check if the distribution of S4 Light is normally distributed
hist(occupancy_estimation$S4_Light, 
     xlab="Light (Lux)", 
     main="Histogram of Sensor 4 Light Values",
     col="skyblue",
     border="black",
     breaks=8)
grid()

par(mfrow=c(1,1))
```

Create a boxpot which shows the Temperature values of each sensor. Use different colors for each sensor.

```{r}
# Create the boxplot
boxplot(occupancy_estimation$S1_Light, occupancy_estimation$S2_Light, occupancy_estimation$S3_Light, 
        occupancy_estimation$S4_Light, 
        names=c("Sensor 1", "Sensor 2", "Sensor 3", "Sensor 4"),
        col=c("green","magenta","orange","red"),
        ylab = "Light (Lux)", main="Boxplot of Light Values by Sensor" )
grid()
```

```{r}
# Consider only days with occupants
light <- subset(occupancy_estimation, Date=="2017-12-22" | Date=="2017-12-23" | Date=="2017-01-10") # Days with occupants

# Create the boxplot
boxplot(light$S1_Light, light$S2_Light, light$S3_Light, light$S4_Light, 
        names=c("Sensor 1", "Sensor 2", "Sensor 3", "Sensor 4"),
        col=c("green","magenta","orange","red"),
        ylab = "Light (Lux)", main="Boxplot of Light Values by Sensor on days with occupants")
grid()
```

```{r}
# Normality test for sensor 4 light values. Sample size must be between 3 and 5000.
shapiro_test <- shapiro.test(light$S4_Light)
print(shapiro_test)

if (shapiro_test$p.value > 0.05) {
  print("Light data for sensor 4 is normally distributed")
} else {
  print("Light data for sensor 4 is not normally distributed")
}
```

Use a histogram to assess normality of the Sound variables.

```{r}
par(mfrow=c(2,2))

# Check if the distribution of S1 Sound is normally distributed
hist(occupancy_estimation$S1_Sound, 
     xlab="Sound (Volts)", 
     main="Histogram of Sensor 1 Sound Values",
     col="skyblue",
     border="black",
     breaks=8)
grid()

# Check if the distribution of S2 Sound is normally distributed
hist(occupancy_estimation$S2_Sound, 
     xlab="Sound (Volts)", 
     main="Histogram of Sensor 2 Sound Values",
     col="skyblue",
     border="black",
     breaks=8)
grid()

# Check if the distribution of S3 Sound is normally distributed
hist(occupancy_estimation$S3_Sound, 
     xlab="Sound (Lux)", 
     main="Histogram of Sensor 3 Sound Values",
     col="skyblue",
     border="black",
     breaks=8)
grid()

# Check if the distribution of S4 Sound is normally distributed
hist(occupancy_estimation$S4_Sound, 
     xlab="Sound (Volts)", 
     main="Histogram of Sensor 4 Sound Values",
     col="skyblue",
     border="black",
     breaks=8)
grid()

par(mfrow=c(1,1))
```
Create a boxpot which shows the Sound values of each sensor. Use different colors for each sensor.

```{r}
# Create the boxplot
boxplot(occupancy_estimation$S1_Sound, occupancy_estimation$S2_Sound, occupancy_estimation$S3_Sound, 
        occupancy_estimation$S4_Sound, 
        names=c("Sensor 1", "Sensor 2", "Sensor 3", "Sensor 4"),
        col=c("green","magenta","orange","red"),
        ylab = "Sound (Volts)", main="Boxplot of Sound Values by Sensor" )
grid()
```
```{r}
# Consider only days with occupants
sound <- subset(occupancy_estimation, Date=="2017-12-22" | Date=="2017-12-23" | Date=="2017-01-10") # Days with occupants

# Create the boxplot
boxplot(sound$S1_Sound, sound$S2_Sound, sound$S3_Sound, sound$S4_Sound, 
        names=c("Sensor 1", "Sensor 2", "Sensor 3", "Sensor 4"),
        col=c("green","magenta","orange","red"),
        ylab = "Sound (Vots)", main="Boxplot of Sound Values by Sensor on days with occupants")
grid()
```

Use a histogram to assess normality of the CO2 variables.

```{r}
par(mfrow=c(1,2))

# Check if the distribution of S5 CO2 is normally distributed
hist(occupancy_estimation$S5_CO2, 
     xlab="CO2 (PPM)", 
     main="Sensor 5 CO2 Values",
     col="skyblue",
     border="black",
     breaks=8)
grid()

# Check if the distribution of S2 CO2 Slope  is normally distributed
hist(occupancy_estimation$S5_CO2_Slope, 
     xlab="CO2 Slope", 
     main="Sensor 5 CO2 Slope Values",
     col="skyblue",
     border="black",
     breaks=8)
grid()

par(mfrow=c(1,1))
```

Create a boxpot which shows the CO2 values.

```{r}
par(mfrow=c(1,2))

# Create the boxplots

boxplot(occupancy_estimation$S5_CO2, 
        names=c("CO2"),
        col=c("green"),
        ylab = "CO2 (PPM)", main="Boxplot of CO2 Values" )
grid()

boxplot(occupancy_estimation$S5_CO2_Slope, 
        names=c("CO2"),
        col=c("green"),
        ylab = "Slope of CO2", main="Boxplot of CO2 Slope Values" )
grid()

par(mfrow=c(1,1))
```
```{r}
# Consider only days with occupants
co2 <- subset(occupancy_estimation, Date=="2017-12-22" | Date=="2017-12-23" | Date=="2017-01-10") # Days with occupants

par(mfrow=c(1,2))

# Create the boxplots

boxplot(co2$S5_CO2, 
        names=c("CO2"),
        col=c("green"),
        ylab = "CO2 (PPM)", main="Boxplot of CO2 Values" )
grid()

boxplot(co2$S5_CO2_Slope, 
        names=c("CO2"),
        col=c("green"),
        ylab = "Slope of CO2", main="Boxplot of CO2 Slope Values" )
grid()

par(mfrow=c(1,1))
```
```{r}
# Normality test for CO2 slope values. Sample size must be between 3 and 5000.
shapiro_test <- shapiro.test(co2$S5_CO2_Slope)
print(shapiro_test)

if (shapiro_test$p.value > 0.05) {
  print("CO2 Slope data is normally distributed")
} else {
  print("CO2 slope data is not normally distributed")
}
```


Create a bar chart of the S6_PIR and S7_PIR variables

```{r}
library(ggplot2)

# Create factors for the S6_PIR and S7_PIR variables
S6_PIR_factor <- factor(occupancy_estimation$S6_PIR, levels=0:1, 
                                      labels=c("No", "Yes"))
S7_PIR_factor <- factor(occupancy_estimation$S7_PIR, levels=0:1, 
                                      labels=c("No", "Yes"))
# Create bar chart of the S6_PIR variable
ggplot(occupancy_estimation, aes(x=S6_PIR_factor)) +
  geom_bar(fill='steelblue') +
  labs(x='Motion Detected?', y='Frequency')

# Create bar chart of the S7_PIR variable
ggplot(occupancy_estimation, aes(x=S7_PIR_factor)) +
  geom_bar(fill='steelblue') +
  labs(x='Motion Detected?', y='Frequency')
```
```{r}
# Consider only days with occupants
motion <- subset(occupancy_estimation, Date=="2017-12-22" | Date=="2017-12-23" | Date=="2017-01-10") # Days with occupants

# Create factors for the S6_PIR and S7_PIR variables
S6_PIR_factor <- factor(motion$S6_PIR, levels=0:1, 
                                      labels=c("No", "Yes"))
S7_PIR_factor <- factor(motion$S7_PIR, levels=0:1, 
                                      labels=c("No", "Yes"))

# Create bar chart of the S6_PIR variable
ggplot(motion, aes(x=S6_PIR_factor)) +
  geom_bar(fill='steelblue') +
  labs(x='Motion Detected?', y='Frequency')

# Create bar chart of the S7_PIR variable
ggplot(motion, aes(x=S7_PIR_factor)) +
  geom_bar(fill='steelblue') +
  labs(x='Motion Detected?', y='Frequency')
```


Create a bar chart of the Room Occupancy Count (target) variable

```{r}
library(ggplot2)

# Create bar chart of Room Occupancy Count
ggplot(occupancy_estimation, aes(x=Occupancy_Count)) +
  geom_bar(fill='steelblue') +
  labs(x='Occupancy Count', y='Frequency')
```
```{r}
# Consider only days with occupants
occupancy <- subset(occupancy_estimation, Date=="2017-12-22" | Date=="2017-12-23" | Date=="2017-01-10") # Days with occupant

# Create bar chart of Room Occupancy Count
ggplot(occupancy, aes(x=Occupancy_Count)) +
  geom_bar(fill='steelblue') +
  labs(x='Occupancy Count', y='Frequency')
```

## Bivariate Analysis

Create a scatter plot matrix of the numeric variables within the data set

```{r}
# Create a scatter plot matrix of the temperature values
df <- c("S1_Temp", "S2_Temp", "S3_Temp", "S4_Temp")
#pairs(occupancy_estimation[df])

# Create a scatter plot matrix of the light values
df <- c("S1_Light", "S2_Light", "S3_Light", "S4_Light")
#pairs(occupancy_estimation[df])

# Create a scatter plot matrix of the sound values
df <- c("S1_Sound", "S2_Sound", "S3_Sound", "S4_Sound")
#pairs(occupancy_estimation[df])

# Create a scatter plot matrix of the CO2 values
df <- c("S5_CO2", "S5_CO2_Slope")
#pairs(occupancy_estimation[df])

# Create a scatter plot matrix of the S1 sensor and CO2 values
df <- c("S1_Temp", "S1_Light", "S1_Sound", "S5_CO2")
pairs(occupancy_estimation[df])

# Create a scatter plot matrix of the S2 sensor and CO2 values
df <- c("S2_Temp", "S2_Light", "S2_Sound", "S5_CO2")
#pairs(occupancy_estimation[df])

# Create a scatter plot matrix of the S2 sensor and CO2 values
df <- c("S3_Temp", "S3_Light", "S3_Sound", "S5_CO2")
#pairs(occupancy_estimation[df])

# Create a scatter plot matrix of the S2 sensor and CO2 values
df <- c("S4_Temp", "S4_Light", "S4_Sound", "S5_CO2")
#pairs(occupancy_estimation[df])
```

Create a boxplot to display the distribution of room occupancy for each day

```{r}
boxplot(occupancy_estimation$Occupancy_Count ~ occupancy_estimation$Date,
        names=c("Dec 22","Dec 23", "Dec 24", "Dec 25", "Dec 26", "Jan 10", "Jan 11"),
        col=c("green"),
        xlab = "Date", ylab = "Occupancy Count", main="Boxplot of Occupancy Counts" )
grid()
```


Create a boxplot to display the distribution of CO2 for each day

```{r}
# Create the boxplot
boxplot(occupancy_estimation$S5_CO2 ~ occupancy_estimation$Date,
        names=c("Dec 22","Dec 23", "Dec 24", "Dec 25", "Dec 26", "Jan 10", "Jan 11"),
        col=c("green"),
        xlab = "Date", ylab = "CO2 (PPM)", main="Boxplot of CO2 Values" )
grid()
```

Create a boxplot to display the distribution of temperature values each day

```{r}
# Create the boxplots

boxplot(occupancy_estimation$S1_Temp ~ occupancy_estimation$Date,
        names=c("Dec 22","Dec 23", "Dec 24", "Dec 25", "Dec 26", "Jan 10", "Jan 11"),
        col=c("green"),
        xlab = "Date", ylab = "Temperature (Celsius)", main="Boxplot of Temperature Values (S1)" )
grid()

boxplot(occupancy_estimation$S2_Temp ~ occupancy_estimation$Date,
        names=c("Dec 22","Dec 23", "Dec 24", "Dec 25", "Dec 26", "Jan 10", "Jan 11"),
        col=c("green"),
        xlab = "Date", ylab = "Temperature (Celsius)", main="Boxplot of Temperature Values (S2)" )
grid()

boxplot(occupancy_estimation$S3_Temp ~ occupancy_estimation$Date,
        names=c("Dec 22","Dec 23", "Dec 24", "Dec 25", "Dec 26", "Jan 10", "Jan 11"),
        col=c("green"),
        xlab = "Date", ylab = "Temperature (Celsius)", main="Boxplot of Temperature Values (S3)" )
grid()

boxplot(occupancy_estimation$S4_Temp ~ occupancy_estimation$Date,
        names=c("Dec 22","Dec 23", "Dec 24", "Dec 25", "Dec 26", "Jan 10", "Jan 11"),
        col=c("green"),
        xlab = "Date", ylab = "Temperature (Celsius)", main="Boxplot of Temperature Values (S4)" )
grid()
```

Create a boxplot to display the distribution of light values each day

```{r}
# Create the boxplots

boxplot(occupancy_estimation$S1_Light ~ occupancy_estimation$Date,
        names=c("Dec 22","Dec 23", "Dec 24", "Dec 25", "Dec 26", "Jan 10", "Jan 11"),
        col=c("green"),
        xlab = "Date", ylab = "Light (Lux)", main="Boxplot of Light Values (S1)" )
grid()

boxplot(occupancy_estimation$S2_Light ~ occupancy_estimation$Date,
        names=c("Dec 22","Dec 23", "Dec 24", "Dec 25", "Dec 26", "Jan 10", "Jan 11"),
        col=c("green"),
        xlab = "Date", ylab = "Light (Lux)", main="Boxplot of Light Values (S2)" )
grid()

boxplot(occupancy_estimation$S3_Light ~ occupancy_estimation$Date,
        names=c("Dec 22","Dec 23", "Dec 24", "Dec 25", "Dec 26", "Jan 10", "Jan 11"),
        col=c("green"),
        xlab = "Date", ylab = "Light (Lux)", main="Boxplot of Light Values (S3)" )
grid()

boxplot(occupancy_estimation$S4_Light ~ occupancy_estimation$Date,
        names=c("Dec 22","Dec 23", "Dec 24", "Dec 25", "Dec 26", "Jan 10", "Jan 11"),
        col=c("green"),
        xlab = "Date", ylab = "Light (Lux)", main="Boxplot of Light Values (S4)" )
grid()
```
Create a boxplot to display the distribution of sound values each day

```{r}
# Create the boxplots

boxplot(occupancy_estimation$S1_Sound ~ occupancy_estimation$Date,
        names=c("Dec 22","Dec 23", "Dec 24", "Dec 25", "Dec 26", "Jan 10", "Jan 11"),
        col=c("green"),
        xlab = "Date", ylab = "Sound (Volts)", main="Boxplot of Sound Values (S1)" )
grid()

boxplot(occupancy_estimation$S2_Sound ~ occupancy_estimation$Date,
        names=c("Dec 22","Dec 23", "Dec 24", "Dec 25", "Dec 26", "Jan 10", "Jan 11"),
        col=c("green"),
        xlab = "Date", ylab = "Sound (Volts)", main="Boxplot of Sound Values (S2)" )
grid()

boxplot(occupancy_estimation$S3_Sound ~ occupancy_estimation$Date,
        names=c("Dec 22","Dec 23", "Dec 24", "Dec 25", "Dec 26", "Jan 10", "Jan 11"),
        col=c("green"),
        xlab = "Date", ylab = "Sound (Volts)", main="Boxplot of Sound Values (S3)" )
grid()

boxplot(occupancy_estimation$S4_Sound ~ occupancy_estimation$Date,
        names=c("Dec 22","Dec 23", "Dec 24", "Dec 25", "Dec 26", "Jan 10", "Jan 11"),
        col=c("green"),
        xlab = "Date", ylab = "Sound (Volts)", main="Boxplot of Sound Values (S4)" )
grid()
```

Display the correlation matrix of the room occupancy and temperature variables of the data set

```{r}
# Extract the room occupancy and temperature variables of the data set
df_temp <- occupancy_estimation[ , c('Occupancy_Count', 'S1_Temp', 'S2_Temp', 'S3_Temp', 'S4_Temp')]

# Display the correlation matrix 
cor(df_temp, method="spearman")
```
Display the correlation matrix of the room occupancy and light variables of the data set

```{r}
# Extract the room occupancy and temperature variables of the data set
df_light <- occupancy_estimation[ , c('Occupancy_Count', 'S1_Light', 'S2_Light', 'S3_Light', 'S4_Light')]

# Display the correlation matrix 
cor(df_light, method="spearman")
```


Display the correlation matrix of the room occupancy and sound variables of the data set

```{r}
# Extract the room occupancy and sound variables of the data set
df_sound <- occupancy_estimation[ , c('Occupancy_Count', 'S1_Sound', 'S2_Sound', 'S3_Sound', 'S4_Sound')]

# Display the correlation matrix 
cor(df_sound, method="spearman")
```

Display the correlation matrix of the room occupancy, CO2, and motion variables of the data set

```{r}
# Extract the room occupancy and CO2 variables of the data set
df_co2 <- occupancy_estimation[ , c('Occupancy_Count','S5_CO2', 'S5_CO2_Slope', 'S6_PIR', 'S7_PIR')]

# Display the correlation matrix 
cor(df_co2, method="spearman")
```

Visualize the correlation plot

```{r}
library("corrplot")

# Extract the room occupancy and sensor data of the data set
df <- occupancy_estimation[ , c('Occupancy_Count', 'S1_Temp', 'S1_Light', 'S1_Sound',
                                'S2_Temp', 'S2_Light', 'S2_Sound',
                                'S3_Temp', 'S3_Light', 'S3_Sound',
                                'S4_Temp', 'S4_Light', 'S4_Sound',
                                'S5_CO2', 'S5_CO2_Slope', 'S6_PIR', 'S7_PIR')]

correlation_matrix <- cor(df, method="spearman")

# Visualize the correlation matrix
corrplot(correlation_matrix, method = 'color', tl.cex = 0.9)
```


## Save for future Exploratory Data Analysis

Use the CO2 variables to build a linear regression model to predict room occupancy.

```{r}
# Find the least-squares line relating room occupancy to CO2 values
model <- lm(Room_Occupancy_Count ~ S5_CO2 + S5_CO2_Slope, data=occupancy_estimation)
summary(model)
```
Use light variables to build a linear regression model to predict room occupancy

```{r}
# Find the least-squares line relating room occupancy to light values
model <- lm(Room_Occupancy_Count ~ S1_Light + S2_Light + S3_Light + S4_Light, data=occupancy_estimation)
summary(model)
```
Use temperature variables to build a linear regression model to predict room occupancy

```{r}
# Find the least-squares line relating room occupancy to temperature values
model <- lm(Room_Occupancy_Count ~ S1_Temp + S2_Temp + S3_Temp + S4_Temp, data=occupancy_estimation)
summary(model)
```
Use sound variables to build a linear regression model to predict room occupancy

```{r}
# Find the least-squares line relating room occupancy to sound values
model <- lm(Room_Occupancy_Count ~ S1_Sound + S2_Sound + S3_Sound + S4_Sound, data=occupancy_estimation)
summary(model)
```
Use motion variables to build a linear regression model to predict room occupancy

```{r}
# Find the least-squares line relating room occupancy to sound values
model <- lm(Room_Occupancy_Count ~ S6_PIR + S7_PIR, data=occupancy_estimation)
summary(model)
```

